Data Migration under a Changing Schema in Ampersand

### 2023-10-18 2022Migration/articleMigrationFACS.tex
sct 4.1: generating migration script needs review by Bas

   Then we define the migration system $\migrsys$ as follows:
   Let $\pair{\dataset}{\schema}$ be the existing system, and let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
\begin{enumerate}
\item We take a disjoint union of the data sets by relabeling relation names:
   \begin{align}
      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
   \end{align}
\item We create transactions to copy equal relations:
   \begin{align}
      \widetilde{\declare{n}{A}{B}}={}&\declare{(\id{copied}.n)}{A}{B}\\
      \begin{split}
         \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
         &\cup\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
      \end{split}
   \end{align}
\item For each new blocking invariant $u$, we generate two helper relations: ${\tt viols}_u$ and ${\tt fixed}_u$, and two corresponding transactions.
   \begin{align}
      \rels_\theenumi={}& \{{\tt viols}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \begin{split}
         \transactions_2={}& \{{\tt viols}_u \mapsfrom \violC{\overrightarrow{u}}\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
         &\cup \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{\overrightarrow{u}}{\dataset'}\mid u \in \rules_{\schema'}-\rules_{\schema}\}
      \end{split}\label{eqn:enforceForRules}
   \end{align}
\item Let us combine the above into a single migration schema, $\schema_\migrsys$:
   \begin{align}
      \concepts_\migrsys ={}& \concepts_\dataset \cup  \concepts_{\dataset'}\\
      \rels_\migrsys ={}& \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'} \cup \rels_3\\
      \rules_\migrsys={}& \emptyset\label{eqn:migrsysrules}\\
      \transactions_\migrsys ={}& \transactions_1\cup\transactions_2
   \end{align}
\end{enumerate}

>>>>

   Then we define the migration system $\migrsys$ as follows:
   Let $\pair{\dataset}{\schema}$ be the existing system, and let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
\begin{enumerate}
\item We take a disjoint union of the data sets by relabeling relation names:
   \begin{align}
      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
   \end{align}
\item We create transactions to copy equal relations:
% \begin{verbatim}
%    RELATION copied_r[A*B]
%    RELATION new_r[A*B] [UNI]
%    ENFORCE copied_r >: new_r /\ old_r
%    ENFORCE new_r >: old_r - copied_r
% \end{verbatim}
   \begin{align}
      \widetilde{\declare{n}{A}{B}}={}&\declare{(\id{copied}.n)}{A}{B}\\
      \begin{split}
         \transactions_1={}&\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}\cap\overleftarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
         &\cup\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
      \end{split}
   \end{align}
\item For each new blocking invariant $u$, we generate a helper relation: ${\tt fixed}_u$, and a transaction to produce its population:
% \begin{verbatim}
%    RELATION fixed_TOTr[A*A]
%    ENFORCE fixed_TOTr >: I /\ new_r;new_r~
%    RULE Block_TOTr : fixed_TOTr |- new_r;new_r~
%    MESSAGE "Relation r[A*B] must remain total."
%    VIOLATION (TXT "Atom ", SRC I, TXT " must remain paired with an atom from B.")

%    ROLE User MAINTAINS TOTr
%    RULE TOTr : I - fixed_TOTr |- new_r;new_r~
%    MESSAGE "Please, make relation r[A*B] total."
%    VIOLATION (TXT "Fix ", SRC I, TXT " by pairing it with an (any) atom from B.")
% \end{verbatim}
   \begin{align}
      \rels_\theenumi={}&\{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \transactions_2={}&\{{\tt fixed}_u\mapsfrom\lambda\dataset'.~\viol{\overleftarrow{u}}{\dataset'}-\viol{\overrightarrow{u}}{\dataset'}\mid u\in\rules_{\schema'}-\rules_{\schema}\}\label{eqn:enforceForRules}
   \end{align}
\item Let us combine the above into a single migration schema, $\schema_\migrsys$:
   \begin{align}
      \concepts_\migrsys ={}& \concepts_\dataset \cup  \concepts_{\dataset'}\\
      \rels_\migrsys ={}& \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'} \cup \rels_3\\
      \rules_\migrsys={}& \emptyset\label{eqn:migrsysrules}\\
      \transactions_\migrsys ={}& \transactions_1\cup\transactions_2
   \end{align}
\end{enumerate}
   After the migration system has been deployed, we must wait until the business has resolved all violations.
   This is something the system must signal, so the migration engineer can switch all traffic to the desired system:
% \begin{verbatim}
   % ROLE User MAINTAINS CleanUp_TOTr
   % RULE CleanUp_TOTr : V[ONE*A] ; (I - fixed_TOTr) ; V[A*ONE]
   % MESSAGE "Now you can remove the migration system because relation r[A*B] is total."
% \end{verbatim}
\begin{equation}
   \forall u\in\rules_{\schema'}-\rules_{\schema}.\ \viol{\overleftarrow{u}}{\dataset'}={\tt fixed}_u
\end{equation}
   After this, the migration system is no longer needed and can be removed.
